Distributed Topology Discovery in Self-Assembled Nano Networks

### 2014-02-18 PAPER/disr.tex
Removed some typos

%------------------------------------------------------------------------------
\section{\disr{} Sketch}
%------------------------------------------------------------------------------

\label{sec:disr_concepts}
The main concept behind the kind of turn prohibition we want to
achieve is the \emph{Segment}. A segment is basically a path of consecutive
nodes and links, starts with a link attached to a node of a different
segment and ending with a link attached to a
node belonging another. In other words, a segment
a path between connecting two other different segments. The example
in Figure~\ref{fig:segments} shows the segment with id $1.4$, starting
with a link connected to the node $9$, the going trought nodes
$4$,$3$,$2$ and ending with the link connected to the node $1$. Other
segments depicted are $9.2$ (link,node $10$,link) and $9.3$
(link from $9$ ,$8$,link,$7$,link,$6$,link,$5$,link to the $1$). Note
also that start/ending link can coincide and a segment can also be
formed by one single link (e.g. in $7.2$ or $5.4$).
An exception to the rule that a segment must connect two other
segments is the first segment established in the network, called
\emph{starting segment} which is a loop beginning/ending on the same
node.

@@

A detailed description of the \disr{} execution model as depicted in
the Figure~\ref{fig:disr_events} will be discussed in
Subsection~\ref{sub:phases}. Roughly speaking, the execution of can be described as the execution
of the following phases: 
\begin{enumerate}
\item Injection of the \disr{} process from an upper layer to set a \emph{bootstrap
node} 
\item Broadcasting and confirmation to create the first segment of the subnet
\item Parallel requests starting from assigned node to discover other
segments. 
\item Turn proibition in each confirmed segment.
\item Draining of remaining packet due timeouts
\end{enumerate}

\section{Detailed Description of \disr{}}
\subsection{Message Types Required}

@@

\begin{figure}
  \centering
    \includegraphics[width=0.25\textwidth]{pictures/dbs_updated.eps}
  \caption{\disr{} node execution model}
  \label{fig:state_machine}
\end{figure}
\begin{itemize}

\begin{figure*}
\centering
\begin{tabular}{cccc}
\includegraphics[width=0.23\textwidth]{pictures/seq01.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq02.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq03.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq04.eps} \\
(a) & (b) & (c) & (d) \\
\includegraphics[width=0.23\textwidth]{pictures/seq05.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq06.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq07.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq08.eps} \\
(e) & (f) & (g) & (h) \\
\end{tabular}
\caption{\disr{} (a) and RPF (b) node coverage}
\label{fig:disr_events}
\end{figure*}

\item{\textbf{Injecting bootstrap request}}: all nodes have an initial
\emph{DBS} status set to \emph{Free}, except for a
 node with status \emph{Bootstrap}, set by some control signal from an
upper layer via $(a)$. When starting, this bootstrap node 
changes its status to \emph{CandidateStarting} $(b)$, injecting a
\texttt{STARTING\_SEGMENT\_REQUEST} across one of its free links
(Figure~\ref{fig:disr_events}-a).

\item{\textbf{Bootnode Broadcasting}}: Each node receiving a \texttt{STARTING\_SEGMENT\_REQUEST},
when $Free$, forwards it to its free links using a flooding mechanism and
becoming $CandidateStarting$ $(g)$.
Each of its free links is then marked as \emph{tvisited} with the segment id
associated to the request. Note that a node that has already received a
\texttt{STARTING\_SEGMENT\_REQUEST} packet can simply ignore further packets
associated to the same request, having already contributed to the
flooding (Figure~\ref{fig:disr_events}-b).
%Note also that each of these packets has a max\_segment\_hops
%field to prevent packets from travelling undefinitively.

@@

%What happens to nodes not receiving these segment request?
%they simply remain \emph{tvisited} and after a timeout reset their
%state as free. But they could also remain \emph{tvisited} for the
%starting segment, since this doesn’t affect their behaviour for future
%segment requests.

@@

\item{\textbf{Injecting Segment Requests}: Each node in the
\emph{Assigned} status can initiate a search for a segment, by sending
a \texttt{SEGMENT\_REQUEST} across one of its free links $(c)$ (see
Figure~\ref{fig:disr_events}-e). 
Note that each \emph{Assigned} note can inject the request just after
becoming assigned, i.e. it is not needed that the whole confirmation
process in completed (Figure~\ref{disr_events}-f).
Note also that nodes previosly set as \emph{CandidateStarting}, when receiving 
a \texttt{SEGMENT\_REQUEST} can simply cancel their
status and set as $Candidate$ for that request, since this new
request means that the first starting segment has already been found $(h)$. 
\item{\textbf{Segment Confirmation}}: The find process is successful
when an already \emph{Assigned} node receives the
\texttt{SEGMENT\_REQUEST} packet. Then, a \texttt{SEGMENT\_CONFIRM} is
sent back along the path that originated the request
(Figure~\ref{fig:disr_events}-g), while the
node remain $Assigned$ (since it could confirm more segments). Each node
previously set as $Candidate$ for that segment id, when receiving the
a confirmation packet changes its status to $Assigned$ $(m)$ and forwards back
the same confirmation until it reaches the initiator of the request,
which then changes from $ActiveSearching$ to $Assigned$ $(d)$.

\item{\textbf{Failing while searching a segment}}: this happen when a node received a \texttt{SEGMENT\_REQUEST} packet but
matches one of two the following conditions: first, the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})$(f)$; the node
is already \emph{Candidate} with another segment id, i.e. another find
process.  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the incoming link (see at
the bottom of Figure~\ref{fig:disr_events}).
%Each of the subsequent nodes will forward this \texttt{SEGMENT\_CANCEL}
%only if no more free links can be explored starting from them.
\end{itemize}

% TODO: FIX and RE-enable in long paper

\section{Detailed Node behaviour Model}
\label{sec:execution_model}

The \emph{LED} variables at each node are are initially set so the node and
each of its links is set as free. The only node that makes exception
is the bootstrap node, initialized in the \emph{Bootstrap} status. All
the subsequent events happening at each node are the consequence of
its current dynamic status (DBS) and the packets received.

%TODO: add state machines

% TODO: fix, move elsewhere or delete
%Assumption: when discussing of visited/\emph{tvisited} nodes, we are assuming
%with an id corresponding to the same subnet. Nodes involved in other
%subnet processes can simply ignore these packets. (TBD: formalize more
%clearly in which cases inter-subnet packets must be ignored, probably
%this happens most of the times, but in some cases, e.g. when a
%terminal node is found, some kind of communication packets are needed
%in order to crearte the bridge between two subnets.)

@@

\subsection{Receiving a \texttt{SEGMENT\_CONFIRM}}
When received, if the node status is \emph{Candidate} with a \emph{segID} corresponding to the one indicated in
the packet, the node learns to belong to the segment associated to the
id. Further, it should forward this packet to the link where the
original \texttt{SEGMENT\_REQUEST} packet came from, so that all candidate nodes
can learn the segment id they belong to.  \emph{LED} shoud be
updated from \emph{tvisited} to \emph{visited}.  the
\emph{link\_tvisited} previously set with the \emph{segID} should be invalidated the
\emph{link\_visited} variable associated to the incoming confirm should be set
as segID. Note that if the request was flooded across different links,
the \emph{link\_visited} variable associated to the other paths should not be
set.  the \emph{link\_visited} variable associated to the link from which the
request was originated should be set segID

\subsection{Receiving a \texttt{SEGMENT\_CANCEL}}
When a node receives a \texttt{SEGMENT\_CANCEL} packet it means that searching
for a segment along that path was unsuccessful. But if the node still
has other free links to try, it should forward a \texttt{SEGMENT\_REQUEST} to
the next link (next according to some internal ordering). So a node
forwards back the \texttt{SEGMENT\_CANCEL} packet along the link that originated
the \texttt{SEGMENT\_REQUEST} packet only when there are no more free links to
try. If this is the case, the node modifies its status from
\emph{Candidate} to \emph{Free} and forwards the \texttt{SEGMENT\_CANCEL} packet to the link from which
the request was received. The process stops when the \texttt{SEGMENT\_CANCEL}
packet reach the starting node that originated the request.

%TODO REPEAT!!
%When a node discovers that the path that includes
%the node itself is no more suitable, a \texttt{SEGMENT\_CANCEL} is sent back
%across the link from which the \texttt{SEGMENT\_REQUEST} was received.  Possible
%reasons for that: The node it’s already candidate for another segment
%The node it’s free but can’t find a free link (according the
%cyclelinks timeout) The node it’s \emph{tvisited} for the same request id,
%previously forwarded that request along some free direction, but now
%it’s receiving a a cancel request and has no more free links
\subsection{Intra-node vs Inter-node parallelism}
In the processes described above, we assumed that each node in the
\emph{Assigned} status can start a find segment process by injecting a
\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when
defining \disr{} should be made in order to decide whether or not the
node must actually perform this action.  The point above is is
strictly related to the general question: which level of “parallelism”
should we allow in \disr{} ? The adopted approach is that, although
the nature of \disr{} itself is intrinsicly parallel, the use of
parallelism makes things work in a more complex way. In other words,
\disr{} is parallel when needed, but it does not exploit parallelism
as an “improving feature”. When not needed, things should be
serialized. For example: when a node is “ready searching” could start
several find segment process associated to the same segment, one for
each free link. But serializing this by investigating the free links
in order could be a simplest solution. We refer to this saying that we
avoid intra-node parallelism.  Note that while intra-node parallelism
can be avoided, trying to avoid inter-node parallelism could be more
complex than allowing the parallelism itself. Imagine for example the
effort of trying to coordinate nodes so that an unique finding node
process is actually running in the whole subnet. Thus, in contrast to
the intra-node parallelism, the inter-node parallelism is a structural
property of the \disr{} algorithm and should not be avoided.
%------------------------------------------------------------------------------

>>>>

%------------------------------------------------------------------------------
\section{\disr{} Overview}
%------------------------------------------------------------------------------

\label{sec:disr_concepts}
The main concept behind the kind of turn prohibition we want to
achieve is the \emph{segment}. A segment $S_y$ is basically a path of consecutive
nodes and links, starting with a link attached to a node belonging to
a different segment $S_x$ and ending with a link attached to a
node of another segment $S_z$. In other words, a segment
a path between connecting two other different segments. The example
in Figure~\ref{fig:segments} shows the segment with id $1.4$, starting
with a link connected to the node $9$, the going through nodes
$4$,$3$,$2$ and ending with the link connected to the node $1$. Other
segments depicted are $9.2$ (link,node $10$,link) and $9.3$
(link from $9$ ,$8$,link,$7$,link,$6$,link,$5$,link to the $1$). Note
also that start/ending link can coincide and a segment can also be
formed by one single link (e.g. in $7.2$ or $5.4$).
An exception to the rule that a segment must connect two other
segments is the first segment established in the network, called
\emph{starting segment} which is a loop beginning/ending on the same
node.

@@

A detailed description of the \disr{} execution model as depicted in
the Figure~\ref{fig:disr_events} will be discussed in
Subsection~\ref{sub:phases}. Roughly speaking, the execution of can be described as the execution
of the following phases: 
\begin{enumerate}
\item Injection of the \disr{} process from an upper layer to set a \emph{bootstrap
node} 
\item Broadcasting and confirmation to create the first segment of the subnet
\item Parallel requests starting from assigned node to discover other
segments. 
\item Turn prohibition in each confirmed segment.
\item Draining of remaining packet due timeouts
\end{enumerate}

\subsection{\disr{} Data Structures}
\label{ssec:disr_data}

We distinct between two different kind of data stored at each node:
\emph{Dynamic Behavior Status} (\emph{DBS}) and \emph{Local
Environment Data} (\emph{LED}).

The \emph{DBS} stores information in order to capture the dynamic
behavior of the node  in the context of the whole
\disr{} execution phased. The \emph{DBS} can have the following values:

\begin{itemize}
\item{\emph{Free}}: a node that has  not been yet considered  by the \disr{} algorithm.
\item{\emph{Bootstrap}}: a node which has been explicitly set as bootstrap node from
an upper layer via. 
\item{\emph{ActiveSearching}}: a node from which a find process of new segments has
been started and not yet cancelled or confirmed. 
\item{\emph{Candidate}}: a node currently candidate for belonging to some segment
, but not being itself the node from which the find process was started. 
\item{\emph{CandidateStarting}}: same as above, but the node is currently being considered as candidate
for starting segment. 
\item{\emph{Assigned}}: a node for which the segment has been determined.  
\end{itemize}

The \emph{LED} are like an instant snapshot of the \disr{} algorithm at each node,
consisting in the following variables:
\begin{itemize}
\item{\emph{segID}}:a value used to specify the segment to which the
node has been assigned or is candidate for being assigned.
\item{\emph{visited}}: a boolean value. When \emph{true}
then a \emph{segID} different from \texttt{NULL} specifies the segment 
to which the node has been assigned. 
\item{\emph{tvisited}}: a boolean value. If \emph{true}, the node is
being considered as candidate for a segment, and the \emph{segID} value
specifies the segment id for which the node is candidate. 
\item{\emph{link\_visited[]}}: an array
of values representing information about attached links, that is, the
segID of the segment owning each link. When \texttt{NULL}, the corresponding link has not yet been
assigned.
\item{\emph{link\_tvisited[]}}: an array of
values representing information about attached links, that is, the \emph{segID} of
the segment for which the link is candidate. When \texttt{NULL}, the link is
not candidate for any segment, i.e. maybe already assigned
or free.
\end{itemize}

\subsection{Message Types Required}

@@

\begin{figure}
  \centering
    \includegraphics[width=0.25\textwidth]{pictures/dbs_updated.eps}
  \caption{\disr{} node execution model}
  \label{fig:state_machine}
\end{figure}

\begin{figure*}
\centering
\begin{tabular}{cccc}
\includegraphics[width=0.23\textwidth]{pictures/seq01.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq02.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq03.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq04.eps} \\
(a) & (b) & (c) & (d) \\
\includegraphics[width=0.23\textwidth]{pictures/seq05.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq06.eps} &
\includegraphics[width=0.23\textwidth]{pictures/seq07.eps} & 
\includegraphics[width=0.23\textwidth]{pictures/seq08.eps} \\
(e) & (f) & (g) & (h) \\
\end{tabular}
\caption{Example of \disr{} segment setup: Nodes status are labeled as CandidateStarting($Cs$), Candidate ($C$), Free ($F$), Assigned ($A$) and Bootstrap ($B$)}
\label{fig:disr_events}
\end{figure*}

\begin{itemize}
\item{\textbf{Injecting bootstrap request}}: all nodes have an initial
\emph{DBS} status set to \emph{Free}, except for a
 node with status \emph{Bootstrap}, set by some control signal from an
upper layer via $(a)$. When starting, this bootstrap node 
changes its status to \emph{CandidateStarting} $(b)$, injecting a
\texttt{STARTING\_SEGMENT\_REQUEST} across one of its free links
(Figure~\ref{fig:disr_events}-a).

\item{\textbf{Bootnode Broadcasting}}: Each node receiving a \texttt{STARTING\_SEGMENT\_REQUEST},
when $Free$, forwards it to its free links using a flooding mechanism and
becoming $CandidateStarting$ $(g)$.
Each of its free links is then marked as \emph{tvisited} with the segment id
associated to the request. Note that a node that has already received a
\texttt{STARTING\_SEGMENT\_REQUEST} packet can simply ignore further packets
associated to the same request, having already contributed to the
flooding (Figure~\ref{fig:disr_events}-b).
%Note also that each of these packets has a max\_segment\_hops
%field to prevent packets from traveling undefinitively.

@@

%What happens to nodes not receiving these segment request?
%they simply remain \emph{tvisited} and after a timeout reset their
%state as free. But they could also remain \emph{tvisited} for the
%starting segment, since this does not affect their behavior for future
%segment requests.

@@

\item{\textbf{Injecting Segment Requests}}: Each node in the
\emph{Assigned} status can initiate a search for a segment, by sending
a \texttt{SEGMENT\_REQUEST} across one of its free links $(c)$ (see
Figure~\ref{fig:disr_events}-e). 
Note that each \emph{Assigned} note can inject the request just after
becoming assigned, i.e. it is not needed that the whole confirmation
process in completed (Figure~\ref{disr_events}-f).
Note also that nodes previously set as \emph{CandidateStarting}, when receiving 
a \texttt{SEGMENT\_REQUEST} can simply cancel their
status and set as $Candidate$ for that request, since this new
request means that the first starting segment has already been found $(h)$. 
\item{\textbf{Segment Confirmation}}: The find process is successful
when an already \emph{Assigned} node receives the
\texttt{SEGMENT\_REQUEST} packet. Then, a \texttt{SEGMENT\_CONFIRM} is
sent back along the path that originated the request
(Figure~\ref{fig:disr_events}-g), while the
node remain $Assigned$ (since it could confirm more segments). Each node
previously set as $Candidate$ for that segment id, when receiving the
a confirmation packet changes its status to $Assigned$ $(m)$ and forwards back
the same confirmation until it reaches the initiator of the request,
which then changes from $ActiveSearching$ to $Assigned$ $(d)$.

\item{\textbf{Failing while searching a segment}}: this happen when a node received a \texttt{SEGMENT\_REQUEST} packet but
matches one of two the following conditions: first, the node is $Free$ but has no
more suitable free links (thus can’t forward the
\texttt{SEGMENT\_REQUEST})$(f)$; the node
is already \emph{Candidate} with another segment id, i.e. another find
process.  In all these cases the node
sends back a \texttt{SEGMENT\_CANCEL} along the incoming link (see at
the bottom of Figure~\ref{fig:disr_events}).
\end{itemize}

\section{Detailed Node behavior Model}
\label{sec:execution_model}

In this Section we provide further details about \disr{}
execution model, describing the internal behavior implemented at
each node. Note that this also corresponds exactly to what should be
implemented from the node-hardware perspective, as discussed in
Section~\ref{sec:implementation}.

The \emph{LED} variables at each node
($segID$,$visited$,$tvisited$,\emph{link\_visited[]},\emph{link\_tvisited[]})
are initially set as \texttt{NULL}, so the node and each of its links
are considered as free.
The only node that makes exception is the bootstrap node, initialized
in the \emph{Bootstrap} status. All the subsequent events happening at
each node are the consequence of its current dynamic status (DBS), the
packets received and the internal status represented by the
\emph{LED}. Note that in the following, when using expression like
``has a free link'' or ``free node'' we will assume the relationship between 
$LED$ and $DBS$ discussed in Section~\ref{sec:disr_data}. Further, for
each packet type all the cases not explicitly listed can simply be
ignored since are associated to invalid/inconsistent events of the
\disr{} execution model. A complete list of these cases can be checked
in the environment simulator source code at~\ref{nanoxim}.

@@

\subsection{Receiving a \texttt{SEGMENT\_CONFIRM}}
When the node status is \emph{Candidate} with a \emph{segID} corresponding to the one indicated in
the packet: the node set itself as $Assigned$ to the segment $SegID$. Further, it should forward this packet to the link where the
original \texttt{SEGMENT\_REQUEST} packet came from, so that all candidate nodes
can learn the segment id they belong to. \emph{LED} should be
updated from \emph{tvisited} to \emph{visited}.

\subsection{Receiving a \texttt{SEGMENT\_CANCEL}}
When a node receives a \texttt{SEGMENT\_CANCEL} packet it means that
searching for a segment along that path was unsuccessful. But if the
node still has some other free link to try, it should forward a
\texttt{SEGMENT\_REQUEST} to the that link. So a node forwards back
the \texttt{SEGMENT\_CANCEL} packet along the link that originated the
\texttt{SEGMENT\_REQUEST} packet $only$ when there are no more free
links to try. If this is the case, the node modifies its status from
\emph{Candidate} to \emph{Free} and forwards the
\texttt{SEGMENT\_CANCEL} packet to the link from which the request was
received. The process stops when the \texttt{SEGMENT\_CANCEL} packet
reach the starting node that originated the request.

\subsection{Intra-node vs Inter-node parallelism}
In the processes described above, we assumed that each node in the
\emph{Assigned} status can start a find segment process by injecting a
\texttt{SEGMENT\_REQUEST}. However, a more accurate decision when
defining \disr{} should be made in order to decide whether or not the
node must actually perform this action.  The point above is is
strictly related to the general question: which level of “parallelism”
should we allow in \disr{} ? The adopted approach is that, although
the nature of \disr{} itself is intrinsically parallel, the use of
parallelism makes things work in a more complex way. In other words,
\disr{} is parallel when needed, but it does not exploit parallelism
as an “improving feature”. When not needed, things should be
serialized. For example: when a node is “ready searching” could start
several find segment process associated to the same segment, one for
each free link. But serializing this by investigating the free links
in order could be a simplest solution. We refer to this saying that we
avoid intra-node parallelism.  Note that while intra-node parallelism
can be avoided, trying to avoid inter-node parallelism could be more
complex than allowing the parallelism itself. Imagine for example the
effort of trying to coordinate nodes so that an unique finding node
process is actually running in the whole subnet. Thus, in contrast to
the intra-node parallelism, the inter-node parallelism is a structural
property of the \disr{} algorithm and should not be avoided.
%------------------------------------------------------------------------------

### 2014-02-18 PAPER/introduction.tex
Removed some typos

Exploring long term alternatives to the CMOS technology is gaining
more and more revelance as the scaling trend of such devices keeps
introducing new challenges: power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{itrs13}. 
While software parallelism and multicore
approaches~\cite{horowitz2004, powell2009} are partially mitigating the impact of such
constraints on performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the Computer Design challenges of the next
decades.

In the last years, Self-assembled nanoscale architectures~\cite{yan2003}
emerged as a promising technology due their tera/peta scale of
integration, defect tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006_1, pistol2009}.
%, patwardhan2006
The main idea behind this approach is to exploit the physical regularity and
stability of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and CNFETs~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing appropriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see Figure~\ref{fig:nana}-a).
A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: \emph{(i) limited node complexity}, \emph{(ii) large scale
randomness} and \emph{(iii) high defect rates}.

The \emph{limited node complexity} aspect is directly
related to the use of complementary DNA tags in order to place circuit
components. In a traditional CMOS process the complexity is introduced
using a photolithography mask, so that larger (more complex) circuits
only require larger masks; conversely, self-assembly achieves
complexity by increasing the number of unique DNA tags, since more
different tags means having more control on component placement. Ideally, by specifying a single
and unique tag for each nano transistor terminal, we could exactly
choose where each component would be placed. But the number of DNA
symbols forming the DNA sequence is limited (sequence of 4 nucleobases
G,A,T,C) and so creating many different tags (of a given lenght) would
mean make them more similar to each other, increasing the probability
of incorrect/partial matching. To avoid this problem we should limit
the number of unique tags, thus limiting the complexity at each node.
In particular, a budget of about $10,000$ CNFETs per node has been
estimated in~\cite{liu_jetcs}.

@@

Finally, \emph{defect rates}: while they are very hardly tolerated in mask
based top down design , the same nature of a bottom up self
assembled process cannot assume such a deterministic device placement
process, thus defect tolerance is something like a design requirement
more than an exception to be avoided.

\begin{figure*}
\centering
\begin{tabular}{cc}
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps} &
    \includegraphics[width=0.40\textwidth]{pictures/dna1_complex2.eps} \\
 (a) & (b)
 \end{tabular}
  \caption{(a) DNA tag matching and (b) resulting trregualr topology}
  \label{fig:nana}
\end{figure*}
These aspects of a DNA-self assembly process lead to some important
implications to be addressed when approaching to the design of
a computer architecture: computational model should be based on a
distributed network of small processing and storage nodes, randomly
placed and interconnected (see Figure~\ref{fig:nana}-b).
Proof-of-concept of such architectures can be found for
example in~\cite{patwardhan2006_1}, where instruction and data
operands are move around the network in order to be processed.
Indepentedly from the particular policy chosen for routing packets,
such networks must be able to guarantee deadlock freedom without any
prior assumption on topology.

In this work we introduce \disr{}, a Distributed Segment-based approach
to deadlock freedom in large scale DNA self assembled networks. Our
contribution aims to achieve the classical properties of a
segment-based approach~\cite{mejia_ipdps06} without requiring any
topology graph, external defect map or centralized algorithm
execution.  So the \disr{} approach is not intended to discover the
``optimal'' segment choice (ideally reachable with the knowledge of
the topology graph) but just to demonstrate a concrete model that can
fit into such complex, irregular and large sized networks.
It is important to
underline that, altougth we can assume an architecture similar to the
one depicted in Figure~\ref{fig:nana} as background, the topology discovery
problem faced in this work is more general and not strictly dependent on any of the
underlying computing architecture.

The paper is organized as follows: in Section~\ref{sec:related_works}
we summarize the main approaches to topology agnostic routing. Next in
Section~\ref{sec:disr_concepts} are described the main elements and
the execution model of the proposed approach.  In Section~\ref{sec:simulation}
simulations are carried out with the opensource tool Nanoxim to
demonstrate how \disr{} preserves some of the main segment-based
properties and how compares to the tree-based approaches. Finally a
draft of a architectural implementation is shown in
Section~\ref{sec:implementation} to evaluate the feasibility of the
approach in terms of node complexity.

\section{Background and Contribution}
\label{sec:related_works}
Several worksaddressed the problem of topology agnostic deadlock
freedom introducing virtual channels~\cite{sancho2002, skeie2002,
skeie2004, koibuchi2003}, but while they show interesting
performances, the low resources deadlock makes them not suitable for
the specific DNA self assembled networks we are assuming as
background.
Other topology independent algorithms~\cite{schroeder1991,
koibuchi2001, cherkasova1996} do not require any virtual channels but
exploit the concept of ``turn prohibition''.
In particular, authors of~\cite{Patwardhan05evaluatingthe} exploits the creation of a spanning tree of the
topology, placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions.
The hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), but this is quite acceptable in classical wide area networks
topologies with a limited number of nodes. Other approaches such as
FX~\cite{sancho2000} mitigate this
issue, but the set of turn restrictions is still prefixed,
strictly depending on the particular tree root selected. 
Other solutions try to approach the issue of irregular
topologies by limiting the number~\cite{duato1997, gomez2004, koibuchi2008} or the
location~\cite{zhang2008, flich2008, liu2011} of missing links. This
restriction is clearly unacceptable given the hight-defect rates of
DNA self-assembled networks scenario. For the same reason, we also
avoid considering solutions based on hardware-redundancy to
dynamically recover defects as in~\cite{constantinides2006, kohler2010,  park2006, ebrahimi2013}.

>>>>

Exploring long term alternatives to the CMOS technology is gaining
more and more relevance as the scaling trend of such devices keeps
introducing new challenges: power density, defect tolerance, testing
costs and wire delays are only a few of the many critical aspects
involved~\cite{itrs13}. 
While software parallelism and multicore
approaches~\cite{horowitz2004} and \cite{powell2009} are partially mitigating the impact of such
constraints on performances, it is likely that the growing computing demand will
eventually need even more radical architectural modifications and new
paradigms in order to address the Computer Design challenges of the next
decades.

In the last years, self-assembled nanoscale architectures~\cite{yan2003}
emerged as a promising technology due their tera/peta scale of
integration, defect tolerance and huge potential computing
capabilities. These technologies are certainly still at
their early stage of development, however different laboratory demos and
prof-of-concepts architectures have been presented~\cite{patwardhan2004, patwardhan2006_1, pistol2009}.
%, patwardhan2006
The main idea behind this approach is exploiting the physical regularity and
stability of DNA structures in order to create a scaffold onto which
nano-devices (e.g. nanowires and CNFETs~\cite{bachtold2001, tans1998, cui2001}) can be
attached. This can be achieved by designing appropriate complementary DNA tags for
each terminal to be placed, so that a nano device will be attached
only where its own DNA tag matches a complementary tag on the DNA grid
scaffold (see Figure~\ref{fig:nana}-a).
A detailed description of the chemical properties involved is far
beyond the scope of this paper (see also~\cite{braun1998, seeman1999}), so we will focus on
the three main challenges that this new fabrication process introduces
in Computer Design: \emph{(i) limited node complexity}, \emph{(ii) large scale
randomness} and \emph{(iii) high defect rates}.

The \emph{limited node complexity} aspect is directly
related to the use of complementary DNA tags in order to place circuit
components. In a traditional CMOS process the complexity is introduced
using a photolithography mask, so that larger (more complex) circuits
only require larger masks; conversely, self-assembly achieves
complexity by increasing the number of unique DNA tags, since more
different tags means having more control on component placement. Ideally, by specifying a single
and unique tag for each nano transistor terminal, we could exactly
choose where each component would be placed. But the number of DNA
symbols forming the DNA sequence is limited (sequence of 4 nucleobases
G,A,T,C) and so creating many different tags (of a given length) would
mean make them more similar to each other, increasing the probability
of incorrect/partial matching. To avoid this problem we should limit
the number of unique tags, thus limiting the complexity at each node.
In particular, a budget of about $10,000$ CNFETs per node has been
estimated in~\cite{liu_jetcs}.

@@

Finally, \emph{defect rates}: while they are very hardly tolerated in mask
based top down design, the same nature of a bottom up self-assembly
 process cannot assume such a deterministic device placement
process, thus defect tolerance is something like a design requirement
more than an exception to be avoided.

\begin{figure*}
\centering
\begin{tabular}{cc}
    \includegraphics[width=0.40\textwidth]{pictures/dna2b.eps} &
    \includegraphics[width=0.40\textwidth]{pictures/dna1_complex2.eps} \\
 (a) & (b)
 \end{tabular}
  \caption{(a) DNA tag matching and (b) resulting irregular topology}
  \label{fig:nana}
\end{figure*}
These aspects of a DNA self-assembly process lead to some important
implications to be addressed when approaching to the design of
a computer architecture: computational model should be based on a
distributed network of small processing and storage nodes, randomly
placed and interconnected (see Figure~\ref{fig:nana}-b).
Proof-of-concept of such architectures can be found for
example in~\cite{patwardhan2006_1}, where instruction and data
operands are move around the network in order to be processed.
Independently from the particular policy chosen for routing packets,
such networks must be able to guarantee deadlock freedom without any
prior assumption on topology.

In this work we introduce \disr{}, a Distributed Segment-based approach
to deadlock freedom in large scale DNA self assembled networks. Our
contribution aims to achieve the classical properties of a
segment-based approach~\cite{mejia_ipdps06} without requiring any
topology graph, external defect map or centralized algorithm
execution.  So the \disr{} approach is not intended to discover the
``optimal'' segment choice (ideally reachable with the knowledge of
the topology graph) but just to demonstrate a concrete model that can
fit into such complex, irregular and large sized networks.
It is important to
underline that, although we can assume an architecture similar to the
one depicted in Figure~\ref{fig:nana} as background, the topology discovery
problem faced in this work is more general and not strictly dependent on any of the
underlying computing architecture.

The paper is organized as follows: in Section~\ref{sec:related_works}
we summarize the main approaches to topology agnostic routing. Next in
Section~\ref{sec:disr_concepts} are described the main elements and
the execution model of the proposed approach.  In Section~\ref{sec:simulation}
simulations are carried out with the open-source tool Nanoxim to
demonstrate how \disr{} preserves some of the main segment-based
properties and how compares to the tree-based approaches. Finally a
draft of a architectural implementation is shown in
Section~\ref{sec:implementation} to evaluate the feasibility of the
approach in terms of node complexity.

\section{Background and Contribution}
\label{sec:related_works}
Several works addressed the problem of topology agnostic deadlock
freedom introducing virtual channels~\cite{sancho2002, skeie2002,
skeie2004, koibuchi2003}, but while they show interesting
performances, the low resources deadlock makes them not suitable for
the specific DNA self assembled networks we are assuming as
background.
Other topology independent algorithms~\cite{schroeder1991,
koibuchi2001, cherkasova1996} do not require any virtual channels but
exploit the concept of ``turn prohibition''.
In particular, authors of~\cite{Patwardhan05evaluatingthe} exploits the creation of a spanning tree of the
topology, placing bidirectional restrictions by avoiding a packet
to traverse the same link in both up and down directions.
The hierarchical nature of this approach can lead to uneven traffic
distribution, with many packets traversing upper links (near to the
root), but this is quite acceptable in classical wide area networks
topologies with a limited number of nodes. Other approaches such as
FX~\cite{sancho2000} mitigate this
issue, but the set of turn restrictions is still prefixed,
strictly depending on the particular tree root selected. 
Other solutions try to approach the issue of irregular
topologies by limiting the number~\cite{duato1997, gomez2004, koibuchi2008} or the
location~\cite{zhang2008, flich2008, liu2011} of missing links. This
restriction is clearly unacceptable given the hight-defect rates of
DNA self-assembled networks scenario. For the same reason, we also
avoid considering solutions based on hardware-redundancy to
dynamically recover defects as in~\cite{constantinides2006, kohler2010,  park2006, ebrahimi2013}.
